\sys: Property-Guided Program Regeneration\\{\vspace{1em}\large Technical University of Denmark, MSc Thesis {\\\vspace{-1em}} Advisors: Christian Gram Kahlauge and Nikos Vasilakis

### 2025-07-18 docs/paper/draft.tex
End-to-end pass
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are brittle and time-consuming.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to fragile prompting strategies.
This work presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis, are checked post hoc to ensure
  correctness, and can be mutated to allow for flexible program transformation through regeneration.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}
\maketitle

\section{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

This work presents \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
As shown in \cref{fig:overview}, given a source program and
a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

\sys enables users to specify transformation goals declaratively.
Logical constraints over properties of the
regenerated program express these goals. Examples are input-output equivalence, side-effect elimination,
target language, lines of code, and others.

Given a program and a query over desired properties, \sys selects a minimal set
of properties to extract from the original program, then attempts to synthesize
a new program that satisfies the query.
\sys casts this process as an optimization problem, where the goal is to find a
program that satisfies the query while preserving as much of the original
program’s properties as possible.
If the query is unsatisfiable under
the constraints or transformations, \sys issues an error
message, showing the conflict.

This section shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

\heading{A compromised javascript library}
The \texttt{flatmap-stream} library was implicated in a high-profile
supply-chain attack that exfiltrated Bitcoin credentials~\cite{ev:eurosec:2022}.
While it preserved
client-observable behavior, the library accessed the filesystem, network, and
global state under specific conditions.
Program regeneration can automatically remove possible malicious payloads
while preserving the original program's observable behavior,
especially in this scenario, where the attack is highly obfuscated and 
activates under precise conditions~\cite{harp:ccs:2021}.
\begin{minted}[frame=lines]{js}
module.exports = function (stream, fn) {
 if (process.env.PRODUCTION) {
  const priv = copay.getKeys()?.[0]?.priv;
  if (priv && wallet.balance > 0) {
    http.request({ hostname: 'evil.com' })
        .send({ priv });
  }} return flatten(steam).apply(fn); };
\end{minted}

\begin{wrapfigure}[3]{r}{.7\columnwidth}
\vspace{-10pt}
\hspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}, \ttt{pure(P)}\}$, where
\texttt{input(P, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.
The, \texttt{pure(P)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.

After 30.4 seconds and one attempt, \sys generates the following program:
\begin{minted}[frame=lines]{js}
module.exports = function(stream, fn) {
  return stream.map(fn); };
\end{minted}

Building on the same source, \sys can perform cross-language
translation. 
Here, the goal is to regenerate a semantically equivalent version
of \texttt{flatmap-stream} in Haskell, to enable integration with a
Haskell-based pipeline or facilitate formal reasoning.

\begin{wrapfigure}[4]{r}{.7\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query asks \sys to preserve I/O behavior and translate the input program to Haskell.
The logic engine deduces that extracting I/O traces and the function signature is
sufficient; the knowledge base automatically associates the target language
\texttt{Haskell} with purity.
\sys transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\texttt{([1,[2,3]], (x)=>x+1)}\to\texttt{[2,3,4]}\rangle$ 
becomes $\langle(\texttt{[1, [2,3]], Js("(x)=>x+1"))}\to\texttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated 20 I/O pairs, \sys produces the following
Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

The following query for \sys uses lines-of-code a coarse metric of
idiomaticity. \sys will regenerate the program, minimizing this metric, while
preserving its I/O behavior.
\begin{wrapfigure}[3]{r}{.7\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \ttt{signature(F)} and \ttt{graph(F)} from
the original program, where \texttt{signature(F)} is the
signature of the function (including its name, return type, and arguments), and \texttt{graph(F)} is the I/O behavior of the function
on a set of generated inputs.
In this case, where the goal is to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until it can make no improvements or there is a negligible difference after two consecutive iterations.
After 15.2 seconds, and two iterations, \sys produces the following program:
\begin{minted}[frame=lines]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number); }
\end{minted}
The regenerated version is easier to verify, portable across compilers, and
avoids undefined behavior.
A user could have also provided performance requirements using 
the \ttt{time(P)} property to further guide the regeneration process.
In this case, \sys would have run the two functions inside a tight loop and 
would have aborted the regeneration process if the performance of the generated
program was not within an acceptable user-defined threshold.

@@

\begin{wrapfigure}[6]{r}{.7\columnwidth}
  \begin{minted}{prolog}
db('music.db').func(f1).func(f2).
trc_sql(p_,I,T) :-
   trc_sql(p,I,T), graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
                   graph(p,I,O).
\end{minted}
\end{wrapfigure}
Having the database initialization being in the same function as 
the database access has negative architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
The query above first specifies \ttt{music.db} as the database to use.
This information is relevant to the \ttt{trc\_sql} plugin, which records the SQL queries 
sent to the databse when the function is driven by an input $i$.
The set of inputs $I$ the function will be driven by is generated by the \ttt{graph} plugin.
The \ttt{call} predicate indicates a function call.
Spefically, the query asks \sys to generate two functions such 
that the graph of the composition of the two is equivalent to the graph (and SQL trace) of the original program.
This transforms each I/O pair to (for example) $\langle\texttt{(f2(f1(), 'Metalica'))}, \texttt{[...]}\rangle$.
\sys first generates 24 inputs, then wraps the node process with monitors that extract the SQL queries.
Then, it executes the original program on each input, recording the output and SQL traces.
Following, \sys transforms the I/O pairs and the SQL traces into a prompt template,
and prompts an LLM instance to generate two functions $f_1$ and $f_2$, which when composed would have the given behavior.
Their composition is then again driven by the same inputs and the SQL traces and outputs are recorded.
After 20.4 seconds and two attempts, \sys arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$ in regards to its graph and SQL trace .
The user could have provided specific signatures and names for each 
function as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function f1() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\heading{Key results}
\sys is able to effectively transform all programs across the above tasks,
producing high-quality outputs that satisfy user-specified constraints.
A total
of six plugins (\ttt{input}, \ttt{graph}, \ttt{signature}, \ttt{pure}, \ttt{trace\_sql}, and \ttt{llm_translate})
were required to perform the above transformations, each
contributing either a set of extracted and verified properties, or a
transformation, and a set of pre- and post-conditions that the logic engine
uses to construct the regeneration plan.

@@

% \sys is a system for program regeneration: given an input program and a set of
% behavioral, structural, or syntactic properties, it synthesizes a new program
% that satisfies a specified subset of those properties. The system treats
% regeneration as a goal-driven process, guided by explicit property
% specifications and a symbolic planner that structures the search space for a
% backend synthesizer.
At a high level, \sys operates in four phases: (1) plan construction, (2)
property extraction, (3) program synthesis, and (4) verification and feedback.
These phases interact through a shared representation of the program and its
transformation goals.

\heading{Plan construction}
First, each of the available plugins contributes a set of pre and post-conditions 
to the logic engine.
Essentially, these describe what needs to hold for an analysis to be applicable to the given 
program and what properties \sys will preserve or transform after regeneration.
Then, the planning component accepts one or more user-provided regeneration goals (termed as the query).
The full set of rules and predicates are used to construct an ASP program
which is then given to an off-the-shelf solver~\cite{gebser2007clasp}.
% TODO: Explain what the solvr does (optimization)
From the solver's output, the planner constructs a plan that specifies which
analyses \sys plugins should apply to the input program and what properties can be
preserved during regeneration in order to fulfill the query.

\heading{Property extraction} \sys then continues by analyzing the input program to
extract the collection of properties.
These may describe the program's interface
(\eg, exported functions), its operational constraints (\eg, disallowing
filesystem access), or its structure (\eg, use of a particular API or programming idiom).
While some properties are derived from generic analyses
applicable across languages~(\eg examples of input-output behavior), others are specific to the input program's language and runtime model.

\heading{Synthesis}
Guided by the transformation plan, \sys creates a prompt template which is then
augmented with the extracted properties, having them normalized to a human-readable text representation.
This prompt is then passed to an LLM instance.
\sys does not include by default the original program's source code in the prompt.
Instead, it uses the extracted properties to guide the LLM instance 
to synthesize a candidate program that satisfies the specified constraints.

\heading{Verification and feedback}
The final phase verifies that the
synthesized program satisfies the target properties.
If verification fails---due to a violated constraint or behavioral divergence---\sys uses this
failure as a signal to refine the plan or re-sample the synthesis step.
This forms a feedback loop: planning, synthesis, and verification interact until \sys
produces a satisfactory program or the system exhausts its search.

@@

\sys represents program characteristics as symbolic \emph{properties}, drawn from a language of first-order predicates.
These properties describe aspects of the input program and serve as regeneration constraints for the output.
They form the basis of both planning and verification.
All aspects of the symbolic representation are expressed as answer-set programming (ASP) facts and rules (an example program is shown in \cref{lst:logic-example})~\cite{Eiter_2009}.

\heading{Answer Set Programming}
ASP \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo} \cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

The syntax, similar to that of \textit{Prolog} and \textit{Datalog}, with code and data being represented
through logical terms. Collections of logical terms, as seen in
\cref{lst:terms}, can represent the game world's state. Rules, expressed using
the \texttt{:-} operator, enable complex reasoning by defining relationships
between atoms. Choice rules, exemplified in \cref{lst:choice}, allow the ASP
solver to make selections among atoms. Integrity constraints, as shown in
\cref{lst:constraint}, can restrict invalid answer sets. Optimization
directives, such as \texttt{\#minimize} and \texttt{\#maximize} in
\cref{lst:optimization}, guide the solver to output optimal answer sets.

>>>>

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are brittle and time-consuming.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to fragile prompting strategies.
This work presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis, are checked post hoc to ensure
  correctness, and can be mutated to allow for flexible program transformation through regeneration.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}
\maketitle

\section{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

This work presents \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
As shown in \cref{fig:overview}, given a source program and
a query, \sys extracts a set of relevant properties, guided by its
logic engine, and synthesizes a new
program that satisfies the given constraints.

@@

Given a program and a query over desired properties,
\sys selects a set
of properties to extract from the original program, then attempts to synthesize
a new program that satisfies the query.
\sys casts this process as an optimization problem, where the goal is to find a
program that satisfies the query.
If the query is unsatisfiable under
the constraints or transformations, \sys issues an error
message, showing the conflict.

This section shows the application of \sys
to four distinct program transformation tasks, 
targeting security,
language translation,
idiomaticity,
and modularity.

\heading{A compromised javascript library}
The \ttt{flatmap-stream} library was implicated in a high-profile
supply-chain attack that exfiltrated Bitcoin credentials~\cite{ev:eurosec:2022}.
While the library preserved
its client-observable behavior, an malicious payload inserted in the module's code by an adversarial developer accessed the filesystem, network, and
global state under specific conditions.
Speficially, 
the library would
(1) detect whether it is running under the dependency tree of a popular bitcoin wallet
application~\cite{copay} and on the live bitcoin network,
(2) collect the user's private keys, 
(3) send those to an attacker-controlled machine,
and finally (4) perform its benign functionality.
Program regeneration can automatically remove such malicious payloads
using the insight that the insight that they retain the original library's functionality,
while performing a side-effectful attack on-the-side~\cite{harp:ccs:2021}.
This approach can secure the component even when the attack is highly obfuscated and 
activates under precise conditions.
The slightly simplified source code of the compromised library is as follows:
\begin{minted}[frame=lines]{js}
module.exports = function (stream, fn) {
 if (process.env.PRODUCTION) {
  const priv = copay.getKeys()?.[0]?.priv;
  if (priv && wallet.balance > 0) {
    http.request({ hostname: 'evil.com' })
        .send({ priv });
  }} return flatten(steam).apply(fn); };
\end{minted}

\begin{wrapfigure}[4]{r}{.7\columnwidth}
\vspace{-10pt}
\hspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
fsignt(p_, S) :- fsignt(p, S).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys query language to express 
this regeneration goal.
This query requires the regenerated program \ttt{p\_} to exhibit the same I/O behavior as the original program \ttt{p}
but forbids side-effects.
\sys takes this query and arrives at the set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\ttt{graph(p, I, O)}, \ttt{inpt(P, I)}, \ttt{fsignt(P, S)}, \ttt{pure(P)}\}$, where
\ttt{input(p, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\ttt{graph(p, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs
(an example I/O pair is $\langle\ttt{([1,[2,3]], (x)=>x+1)}\to\ttt{[2,3,4]}\rangle$);
and \ttt{fsignt(p, S)} extracts the function's signature using a language-aware parser.
The, \ttt{pure(p)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.
After 30.4 seconds and one attempt, having generated 30 I/O pairs, \sys generates the following program:
\begin{minted}[frame=lines]{js}
module.exports = function(stream, fn) {
  return stream.map(fn); };
\end{minted}

Building on the same source, \sys can perform cross-language
translation. 
Here, the goal is to regenerate a semantically equivalent version
of \ttt{flatmap-stream} in Haskell, to enable integration with a
Haskell-based pipeline or facilitate formal reasoning.

\begin{wrapfigure}[7]{r}{.5\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, Ih, Oh) 
  :- graph(p, Ijs, Ojs),
     tr(Ijs, Ih),
     tr(Ojs, Oh).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query asks \sys to preserve I/O behavior and translate the input program to Haskell.
The logic engine deduces that extracting I/O traces and the function signature is
sufficient.
The \ttt{tr} predicate is a generic LLM-assisted translator, which 
means that an LLM instance will be used to convert the inputs and outputs from Javascript 
to Haskell.
In this case, the given query is sufficient for the translation to succeed, 
but a natural-language description of the transformation can be passed as an additional 
argument to \ttt{tr}.
\sys, same as before, extracts an I/O trace from the original program.
Then, it transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\ttt{([1,[2,3]], (x)=>x+1)}\to\ttt{[2,3,4]}\rangle$ 
becomes $\langle(\ttt{[1, [2,3]], Js("(x)=>x+1"))}\to\ttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated and subsequently translated 20 I/O pairs,
\sys produces the following Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

The following query for \sys uses lines-of-code a coarse metric of
idiomaticity. \sys will regenerate the program, minimizing this metric, while
preserving its I/O behavior.
\begin{wrapfigure}[3]{r}{.7\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \ttt{signature(F)} and \ttt{graph(F)} from
the original program, where \ttt{signature(F)} is the
signature of the function (including its name, return type, and arguments), and \ttt{graph(F)} is the I/O behavior of the function
on a set of generated inputs.
In this case, where the goal is to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until it can make no improvements or there is a negligible difference after two consecutive iterations.
After 15.2 seconds, and two iterations, \sys produces the following program:
\begin{minted}[frame=lines]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number); }
\end{minted}
The regenerated version is easier to verify, portable across compilers, and
avoids undefined behavior.
A user could have also provided performance requirements using 
the \ttt{time(P)} property to further guide the regeneration process.
In this case, \sys would have run the two functions inside a tight loop and 
would have aborted the regeneration process if the performance of the generated
program was not within an acceptable user-defined threshold.

@@

\begin{wrapfigure}[7]{r}{.7\columnwidth}
  \begin{minted}{prolog}
db('music.db').func(f1).func(f2).
trc_sql(p_,I,T) :-
   trc_sql(p,I,T), graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
                   graph(p,I,O).
\end{minted}
\end{wrapfigure}
Having the database initialization being in the same function as 
the database access has negative architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
The query above first specifies \ttt{music.db} as the database to use.
This information is relevant to the \ttt{trc\_sql} plugin, which records the SQL queries 
sent to the databse when the function is driven by an input $i$.
The set of inputs $I$ the function will be driven by is generated by the \ttt{graph} plugin.
The \ttt{call} predicate indicates a function call.
Spefically, the query asks \sys to generate two functions such 
that the graph of the composition of the two is equivalent to the graph (and SQL trace) of the original program.
This transforms each I/O pair to (for example) $\langle\ttt{(f2(f1(), 'Metalica'))}, \ttt{[...]}\rangle$.
\sys first generates 24 inputs, then wraps the node process with monitors that extract the SQL queries.
Then, it executes the original program on each input, recording the output and SQL traces.
Following, \sys transforms the I/O pairs and the SQL traces into a prompt template,
and prompts an LLM instance to generate two functions $f_1$ and $f_2$, which when composed would have the given behavior.
Their composition is then again driven by the same inputs and the SQL traces and outputs are recorded.
After 20.4 seconds and two attempts, \sys arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$ in regards to its graph and SQL trace .
The user could have provided specific signatures and names for each 
function as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function f1() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\heading{Key results}
\sys is able to be effectively applied to all the above transformation goals,
producing high-quality outputs that satisfy user-specified constraints.
A total
of six plugins (\ttt{input}, \ttt{graph}, \ttt{fsignt}, \ttt{pure}, \ttt{trace\_sql}, and \ttt{tr})
were required to perform the above transformations, each
contributing either a set of extracted and verified properties, or a
transformation, and a set of pre- and post-conditions that the logic engine
uses to construct the regeneration plan.

@@

% \sys is a system for program regeneration: given an input program and a set of
% behavioral, structural, or syntactic properties, it synthesizes a new program
% that satisfies a specified subset of those properties. The system treats
% regeneration as a goal-driven process, guided by explicit property
% specifications and a symbolic planner that structures the search space for a
% backend synthesizer.
At a high level, \sys operates in four phases: (1) plan construction, (2)
property extraction, (3) program synthesis, and (4) verification and feedback.
These phases interact through a shared representation of the program.

\heading{Plan construction}
First, each of the available plugins contributes a set of pre and post-conditions 
to the logic engine.
These describe what needs to hold for an analysis to be applicable to the given 
program and what properties \sys will preserve or transform after regeneration.
Then, the planning component accepts one or more user-provided regeneration goals (termed as the query).
The full set of rules and predicates are used to construct a logic program
which is then given to an off-the-shelf solver~\cite{gebser2007clasp}.
From the solver's output, the planner constructs a plan that specifies which
analyses \sys plugins should apply to the input program and what properties can be
preserved during regeneration in order to fulfill the query.
If the query is unsatisfiable, \sys halts and produces an error message.

\heading{Property extraction} \sys then continues by analyzing the input program to
extract the collection of properties.
These may describe the program's interface
(\eg, exported functions), its operational constraints (\eg, disallowing
filesystem access), or its structure (\eg, use of a particular API or programming idiom).
Some properties are derived from generic analyses
applicable across languages~(\eg input-output behavior) 
and others are specific to the input program's language and runtime model.
Plugins insert preconditions for the analyses they provide into the \sys knowledge base.

\heading{Synthesis}
Guided by the transformation plan, \sys creates a prompt template which it then
augments with the extracted properties, having them normalized to a human-readable text representation.
This prompt is then passed to an LLM instance.

\heading{Verification and feedback}
The final phase verifies that the
synthesized program satisfies the target properties.
If verification fails---due to a violated constraint---\sys uses this
failure as a signal to refine the plan or re-sample the synthesis step.
This forms a feedback loop: planning, synthesis, and verification interact until \sys
produces a satisfactory program or the system exhausts its search.

@@

\sys represents program characteristics as symbolic \emph{properties}, drawn from a language of first-order predicates.
These properties describe aspects of the input program and serve as regeneration constraints for the output.
They form the basis of both planning and verification.
All aspects of the symbolic representation are expressed as logic facts and rules (an example program is shown in \cref{lst:logic-example}).

\begin{listing}[t]
  \begin{minted}[frame=lines, fontsize=\small]{prolog}
% --- Base program facts ---
language(p, c).

% --- Plugin: len ---
can(len(p)). 
% Plugin: signature ---
can(signature(p)).
% ---  Plugin: tr ---
can(tr(p, haskell)).
  do(language(p, haskell)) :- do(tr(p, haskell)).
% ---- Axioms ----
:- X, not can(X). % All transformations must be possible
:- language(p, L1), language(p, L2)),
    L1 != L2. % A program can have one language

% --- User query ---
#min len(p). % User query
\end{minted}
  \caption{\textbf{Example \sys logic program.}
  This is a lightly simplified version of the logic program that produces the plan for the idiomatization task in \cref{sec:example}. 
  The program contains facts coming from (1) the input program (written in C),
  (2) three plugins (\ttt{len}, \ttt{signature}, and \ttt{tr}),
  (3) axioms from the logic engine (\eg a program can have one language, an analysis will be applied only if possible, all goals must be satisfied), and
  (4) the user query (the user wants to minimize the program's length).
  }
  \label{lst:logic-example}
\end{listing}

\heading{Property predicates}
Each property is a logical predicate over the program symbol $p$,
or a derivative symbol extracted from $p$.
For example, 
the \ttt{language}$(p, \ttt{javascript})$ predicate describes that the input program is written in JavaScript.
The \ttt{fsignt}$(p, S)$ predicate is used to extract well-formed function signature $S$ when the program \ttt{p} exposes one.
If the user wanted to refer to each function in the program, they could 
combine the predicates \ttt{func(F, p)} and \ttt{fsignt(F, S)} in a single rule in their query.
The \ttt{pure}$(p)$ predicate can be used as a contraint
to signify that the regenerate program
should not have side-effects (\eg file system access, network calls, \etc).

A property $X$ appears as either an
  plugin-provided $\mathsf{can}(X)$ atom, which denotes that the system is capable of extracting or inferring property $X$ from the input.
  A pure $X$ atom denotes that either the user has explicitly requested property $X$ to be enforced, or the planner has deduced that it is necessary to satisfy the query.

\sys also supports minimization, maximization and elimination of properties
through the $\mathsf{min}$, $\mathsf{max}$, and $\mathsf{not}$ directives.

\heading{Planning over properties}
The logic engine determines the set of active properties to enforce by solving
a constrained optimization problem over the extractable property set $P_r$.
It satisfies the
following key constraint:
\begin{align*}
  X &\Leftarrow \mathsf{can}(X) &\text{Apply only extractable properties}
\end{align*}

\heading{Plugin-defined rules}
Plugins can also contribute domain-specific inference rules and constraints.
For example, a plugin for JavaScript-to-Haskell translation may define:
\begin{align*}
\mathsf{can}(\ttt{translate}_{\ttt{js} \rightarrow \ttt{hs}}(p)) & \\
\ttt{translate}_{\ttt{js} \rightarrow \ttt{hs}}(p) &\Rightarrow \ttt{language}(p, \ttt{haskell}) \\
\ttt{translate}_{\ttt{js} \rightarrow \ttt{hs}}(p) &\Rightarrow \mathsf{language}(p, \ttt{javascript})
\end{align*}
Intuitively, these rules mean (1) that the system is able to apply the plugin,
and (2) it can only be applied if the original program is written in JavaScript, 
and the regenerated program \emph{must} be written in Haskell.

\heading{Axioms and domain knowledge}
The logic engine also includes domain-wide axioms based on some common assumptions.
For example
$\ttt{language}(p, \ttt{haskell}) \Rightarrow \ttt{pure}(p)$
captures the assumption that programs written in Haskell are pure by default.
Such axioms allow inferred properties to trigger downstream constraints or synthesis behaviors.

\begin{figure}[t]
\centering
\small
\begin{tabular}{l}
\textbf{toasp} (q : Query) : ASPCode {$\leftarrow$} \textbf{match} q \textbf{with} \\[0.5ex]
\quad \textbar~$X$: ungroundable($X$) {$\rightarrow$} \ttt{do(}$X$\ttt{)} \\
\quad \textbar~\#\ttt{min}~$X$ {$\rightarrow$} \ttt{do\_min(}$X$\ttt{)} \\
\quad \textbar~\#\ttt{max}~$X$ {$\rightarrow$} \ttt{do\_max(}$X$\ttt{)} \\
\textbf{end}
\end{tabular}
\caption{DSL-to-ASP compilation rules. Ungroundable atoms are wrapped in \ttt{do} to denote candidate actions. Optimization directives are compiled to \ttt{do\_min} and \ttt{do\_max}.}
\label{fig:asp-compilation}
\end{figure}

In the logic space, properties are encoded as predicates over abstract program
identifiers.
To bridge this abstraction gap, \sys given the set of true predicates
from the logic space, extracts the corresponding concrete properties from the
source program.
For example, the predicate $\ttt{inputs}(p, I)$ in the logic space
will be concertized to set of inputs $I$ by invoking the \ttt{input} plugin,
which uses an LLM to generate a set of inputs for the program $p$.
Similarly, the predicate $\ttt{graph}(p, I, O)$ will be concertized to a set of
input-output pairs $(I, O)$ by executing the program $p$ on the inputs $I$ and
recording the outputs $O$.
After regeneration, the predicates over the concrete regenerated program $p\_$
will be checked regarding their relationship to the same properties over program $p$
in the logic space.
An equality relationship is one where the predicate's grounded version
$X$ is the same between the two programs, modulo the program identifier~(\eg, $\ttt{graph}(p, i, o) \equiv \ttt{graph}(p\_, i, o)$).
Equality is defined differently for each property, and is enforced by the corresponding plugin.
Similarly, minimization and maximization relationships are defined as
the concrete property value being less than or greater than the original program's value, respectively.
Again, the less-than and greater-than operators are defined by the plugin that verifies the property.

As shown in \cref{fig:logic-to-concrete}, this process constitutes a
translation barrier: declarative goals over properties like \ttt{inputs},
\ttt{outputs}, and \ttt{traces} are realized through observation and
instrumentation. Once aligned, the logic engine can reason over these grounded
facts to drive transformation.

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1JdOfgpaVUy7_EgFyguRXHxFKdmZwyHaiHshnifHG0TQ/edit?usp=sharing
  \includegraphics[width=.9\columnwidth]{figs/kvasir_logic-space.pdf}
  \caption{\textbf{\sys transforms a program over its properties from the abstract logic space to the concrete space.}
  Relationships between original and regenerated program's properties are inferred 
  from the query in combination with the knowledge base and enforced by \sys plugins.
  }
  \label{fig:logic-to-concrete}
\end{figure}

\heading{Extensibility}
\sys decentralizes property extraction: plugins contribute $\mathsf{can}(\cdot)$ facts, and pre- and post-conditions, while the core planner resolves which properties to enforce based on goals, capabilities, and axioms.
This allows many different types of analyses (language-aware, language-agnostic, or LLM-assisted)
to coexist in a unified planning layer.

\heading{Answer Set Programming}
The \sys prototype compiles each input query to an answer-set programming (ASP) program~\cite{Eiter_2009}.
A brief introduction to the paradigm follows.
ASP \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo} \cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

The syntax is similar to \textit{Prolog} and \textit{Datalog}, with code and data being represented
through logical terms.
Collections of logical terms, as seen in
\cref{lst:terms}, can represent the program and its properties.
Rules, expressed using
the \ttt{:-} operator, enable complex reasoning by defining relationships
between atoms.
Choice rules, exemplified in \cref{lst:choice}, allow the ASP
solver to make selections among atoms.
Integrity constraints, as shown in
\cref{lst:constraint}, can restrict invalid answer sets. Optimization
directives, such as \ttt{\#minimize} and \ttt{\#maximize} in
\cref{lst:optimization}, guide the solver to output optimal answer sets,
with respect to a given metric (\eg the number of properties to be extracted from the input program).
